---
title: 'Vignette for equivalence class-level analyses'
author:
- name: Jeroen Gilis
  affiliation: Ghent University, Ghent, Belgium
bibliography: satuRn.bib
date: "16/02/2023"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
  BiocStyle::pdf_document: default
package: satuRn
abstract: |
  Vignette that describes how to use satuRn for performing differential usage 
  analyses on equivalence classes.
vignette: >
  %\VignetteIndexEntry{Vignette for equivalence class-level analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, we demonstrate how to perform a **differential equivalence**
**class (DECU) analysis** with `satuRn`, as opposed to the more common 
differential transcript usage analysis. We will work on a subset of the 
single-cell transcriptomics (10X Genomics Chromium) dataset from Hagai *et al.* 
For a more detailed description on the biological context of the experiment, we 
refer to the paragraph `Import equivalence classes` of this vignette and the 
original publication. 

The dataset by Hagai *et al.* was generated using the 10X Genomics Chromium 
sequencing protocol. This protocol generates 75 basepair, paired-end reads, one 
of which harbors the unique molecular identifier (UMI). As a consequence, only 
one read is used for mapping the against a reference transcriptome or genome. 
This is typically considered insufficient for obtaining transcript-level
quantifications. 

However, it is possible to still obtain sub gene-level counts by quantifying 
equivalence classes. An equivalence class is the set of transcripts that a 
sequencing read is compatible with. As such, reads that are aligned to the same 
set of transcripts are part of the same equivalence class. A transcript 
compatibility count is the number of reads assigned to each of the different 
equivalence classes. For a more detailed biological interpretation of an 
equivalence class, we refer to the `visualize DTU` paragraph in this vignette, 
or to our `satuRn` publication. Note that we are by no means the first to 
suggest performing differential expression analyses on equivalence classes; see
ADD REFERENCES for relevant examples.

Note that for `satuRn` there is no real distinction between performing a 
transcript-level or equivalence class-level analysis. Once a proper input object
is provided, with each row corresponding to a sub gene-level feature, `satuRn` 
will perform a differential usage analysis regardless of the specific feature 
type.

# Load and wrangle data

## Load libraries

```{r, message=FALSE, warning=FALSE}
library(tximportData)
library(SingleCellExperiment)
library(satuRn)
library(ggplot2)
library(edgeR)
library(limma)
library(fishpond)
```

## Import equivalence classes

For this vignette, we will perform a toy analysis on a subset of the dataset
from Hagai *et al.* The authors performed bulk and single-cell transcriptomics 
experiments that aimed to characterize the innate immune response’s 
transcriptional divergence between different species and the variability in 
expression among cells. Here, we focus on a small subset of one of the 
single-cell experiments (10X Genomics sequencing protocol), which characterized 
mononuclear phagocytes in mice under two treatment regimes: unstimulated (unst) 
and after two hours of stimulation with lipopolysaccharide (LPS2). To reduce the
runtime for this vignette, we only quantified the expression in 50 cells of each
treatment group, based on the original whitelist (essentially a list of high 
quality cells as determined by the CellRanger software that is often used to
analyze 10X Genomics data) provided by the original authors.

To obtain equivalence class-level counts, we quantified the fastq files from 
Hagai *et al.* using the Alevin software. Alevin can take a predetermined 
whitelist of high quality cells as input, which allows for quantifying only a 
subset of the cells. More importantly, Alevin allows for directly obtaining 
equivalence class-level counts by using the `-dumpBfh` flag. For each data 
sample, the equivalence-class counts are outputted in a raw format in a 
`bfh.txt` file. These files can we read and wrangled into a R-friendly format
using the `alevinEC` function of the `fishpond` R package.

Note that other pseudo-alignment tools such as `Salmon` (which is internally 
used by Alevin) and `kallisto` are also able to output equivalence class counts.

Here, we read and wrangle the equivalence class counts from the `bfh.txt` files
of the subset of samples from the dataset by Hagai *et al.* that is included in 
the `tximportData` package.

```{r}
# set path to the bfh.txt files generated by Alevin with the `-dumpBfh` flag
# that are present in the `tximportData` package
dir <- system.file("extdata", package="tximportData")
files <- c(file.path(dir,"alevin/mouse1_unst_50/alevin/bfh.txt"),
           file.path(dir,"alevin/mouse1_LPS2_50/alevin/bfh.txt"))

# Additionally read a dataframe that links transcript identifiers to gene 
# identifiers
tx2gene <- read.table(file.path(dir, "tx2gene_alevin.tsv"),
                      sep = "\t",
                      header = FALSE)
colnames(tx2gene) <- c("isoform_id", "gene_id")

# use `alevinEC` to read and wrangle the equivalence class counts (~10 seconds)
EC_list <- alevinEC(paths = files, 
                    tx2gene = tx2gene, 
                    multigene = FALSE,
                    ignoreTxVersion = FALSE,
                    ignoreAfterBar = FALSE,
                    quiet = TRUE)
```

The `AlevinEC` function outputs a list. The *counts* element of the list is
a sparse matrix, where each row corresponds to an equivalence class and each
column corresponds to a cell identifier. The *_integer* at the end of the cell
identifier corresponds to the sample (more specifically, the index of the 
element in the *files* vector) from which the cell originates.

```{r}
TCC_mat <- EC_list$counts
dim(TCC_mat)
TCC_mat[1:3,1:3]
```

Note that the equivalence class names are in the form of integers separated by
"|" signs. These integers allow for linking back equivalence classes to their
corresponding genes and gene isoforms (transcripts) by using the 
*tx2gene_matched* element of the `AlevinEC` output. For example the first
equivalence class, *2183|2187|2188|2190* corresponds to the genes and 
transcripts in rows 2183, 2187, 2188 and 2190 of the *tx2gene_matched* element.

```{r}
tx2gene_matched <- EC_list$tx2gene_matched
dim(tx2gene_matched)
head(tx2gene_matched)

tx2gene_matched[c(2183,2187,2188,2190),]
```

## Construct feature-level information

Next, we construct a dataframe that allows for linking each equivalence class
(feature) to its corresponding gene and transcript(s). This can be achieved 
using the following code.

```{r}
eccs <- strsplit(rownames(TCC_mat),"\\|",fixed=FALSE)

# link ECs to transcripts
txForEachEC <- lapply(eccs, function(ecc){
  tx2gene_matched$isoform_id[as.integer(ecc)]
})

# link ECs to genes
eccs <- sapply(eccs, function(x) x[1]) # we can simply take the first transcript
# identifier, since we already made sure all transcripts in the equivalence 
# class come from the same gene by by setting multigene = FALSE in the alevin_EC
# function. See the help files of AlevinEC for more information on the function
# arguments
geneForEachEC <- tx2gene_matched$gene_id[as.integer(eccs)]

rd <- data.frame(isoform_id = rownames(TCC_mat),
                 gene_id = geneForEachEC,
                 row.names = rownames(TCC_mat))
rd$tx_id <- txForEachEC
head(rd)
```

In the `isoform_id` column, we have a copy of the equivalence class names. Note 
that the target features (here the equivalence class) must always be in a column
named `isoform_id` for downstream analysis with `satuRn`. 

In the `gene_id` column, we have the name of the gene to which the equivalence
class corresponds. Note that because we imported the data with the `AlevinEC`
function with the `multigene` option set to FALSE. As such, each equivalence 
class will be a set of reads that is only compatible with a single gene. In the
context of a differential usage analysis, this is the most sensible thing to do;
if we would retain reads compatible with multiple genes, it would not be 
possible to compute their usage within their gene (this would result in multiple
usages per feature). Hence, we remove such classes, which in our hands removes
approximately 15% of the total reads, which is in line with previous reports.

Finally, the `tx_id` column displays the set of transcripts with which each
equivalence class is compatible. This column is useful for downstream biological
interpretation of the equivalence classes.

## Construct cell-level information 

Here we provide the cell-level information. Cells originated either from the
unstimulated or LPS2 treatment group.

```{r}
cd <- data.frame(cellID = colnames(EC_list$counts),
                 mouse = factor(rep("mouse_1",times=100)),
                 treatment = factor(rep(c("unst", "LPS2"),each=50),
                                    levels = c("unst", "LPS2")),
                 row.names = colnames(EC_list$counts))
head(cd)
```

## Construct SingleCellExperiment object

We combine the counts, feature-level information and cell-level information into
a `SingleCellExperiment` object.

```{r}
sce <- SingleCellExperiment(assays = list(counts = TCC_mat), 
                            colData = cd, 
                            rowData = rd)
sce
```

## Feature-level filtering

Here we perform some feature-level filtering. For this task, we adopt 
the filtering criterion that is implemented in the R package `edgeR`. 
Alternatively, one could adopt the `dmFilter` criterion from the `DRIMSeq` R 
package, which provides a more stringent filtering when both methods are run 
in default settings. After filtering, we again remove transcripts that are 
the only isoform expressed of a certain gene.

```{r}
# Remove lowly expressed equivalence classes
filter_edgeR <- filterByExpr(assay(sce),
                             min.count = 1,
                             min.total.count = 0,
                             large.n = 20,
                             min.prop = 0)
table(filter_edgeR)

sce <- sce[filter_edgeR,]

# Remove EC that are the only EC expressed of a certain gene
uniqueGene <- which(isUnique(rowData(sce)$gene_id))
sce <- sce[-uniqueGene,]

dim(sce)
```

# satuRn analysis

We here perform a canonical `satuRn` analysis with equivalence classes as 
feature type. For amore elaborate description of the different steps, we refer 
to the main 
[vignette](https://www.bioconductor.org/packages/devel/bioc/vignettes/satuRn/inst/doc/Vignette.html) 
of the `satuRn` package.

## Fit quasi-binomial generalized linear models

```{r}
sce <- satuRn::fitDTU(
    object = sce,
    formula = ~ 0 + treatment,
    parallel = FALSE,
    BPPARAM = BiocParallel::bpparam(),
    verbose = TRUE
)
```

## Test for differential equivalence class usage

Here we test for differential equivalence class usage between the two treatment
groups.

### Set up contrast matrix

First, we set up a contrast matrix, which specifies which combination of model
parameters corresponds to our research question, i.e., testing differential 
usage between the treatment groups. This can be done automatically with the 
`makeContrasts` function of the `limma` R package.

```{r}
group <- sce$treatment
design <- model.matrix(~ 0 + group) # construct design matrix
colnames(design) <- levels(group)

L <- limma::makeContrasts(
    Contrast1 = unst - LPS2,
    levels = design)
L # contrast matrix
```

### Perform the test

Next we can perform differential usage testing using `testDTU`. We again adopt 
default settings. For more information on the parameter settings, please consult
the help file of the `testDTU` function.

```{r}
sce <- satuRn::testDTU(
    object = sce,
    contrasts = L,
    diagplot1 = TRUE,
    diagplot2 = TRUE,
    sort = FALSE,
    forceEmpirical = FALSE
)
```

When set to TRUE, the `diagplot1` and `diagplot2` arguments generate diagnostic 
plots. See our main 
[vignette](https://www.bioconductor.org/packages/devel/bioc/vignettes/satuRn/inst/doc/Vignette.html) 
for a detailed explanation on their interpretation.

The test results are now saved into the `rowData` of our SummarizedExperiment 
object under the name `fitDTUResult_` followed by the name of the contrast 
of interest (i.e. the column names of the contrast matrix). 
The results can be accessed as follows:

```{r}
head(rowData(sce)[["fitDTUResult_Contrast1"]]) # first and only contrast
```

See our main 
[vignette](https://www.bioconductor.org/packages/devel/bioc/vignettes/satuRn/inst/doc/Vignette.html) 
for a detailed explanation on the interpretation of the different columns.

## Visualize DTU

### Top 3 features

Finally, we may visualize the usage of the top 3 differentially used equivalence
classes in the different treatment groups. By the setting the `transcripts` and
`genes` arguments to `NULL` and specifying `top.n =3`, the 3 features with the
smallest (empirically correct) false discovery rates are displayed.

```{r, warning=FALSE}
group1 <- colnames(sce)[sce$treatment == "unst"]
group2 <- colnames(sce)[sce$treatment == "LPS2"]

plots <- satuRn::plotDTU(
    object = sce,
    contrast = "Contrast1",
    groups = list(group1, group2),
    coefficients = list(c(1, 0), c(0, 1)),
    summaryStat = "model",
    transcripts = NULL,
    genes = NULL,
    top.n = 3)

# to have same layout as in our publication
for (i in seq_along(plots)) {
    current_plot <- plots[[i]] +
        scale_fill_manual(labels = c("unst", "LPS2"), 
                          values = c("royalblue4","firebrick")) +
        scale_x_discrete(labels = c("unst", "LPS2"))
    print(current_plot)
}
```

Below, we discuss how differential usage between treatment groups of equivalence
classes can be interpreted biologically. We discuss this for two genes as 
examples.

### Biological interpretation for gene ENSMUSG00000030560.18

We here plot the differential usage of all the equivalence classes (retained
after feature-level filtering) of the gene ENSMUSG00000030560.18. We plot
each equivalence class of this gene by specifying 
`genes = "ENSMUSG00000030560.18"`.

```{r}
plots <- satuRn::plotDTU(
    object = sce,
    contrast = "Contrast1",
    groups = list(group1, group2),
    coefficients = list(c(1, 0), c(0, 1)),
    summaryStat = "model",
    transcripts = NULL,
    genes = "ENSMUSG00000030560.18",
    top.n = 3)

# to have same layout as in our publication
for (i in seq_along(plots)) {
    current_plot <- plots[[i]] +
        scale_fill_manual(labels = c("unst", "LPS2"), 
                          values = c("royalblue4","firebrick")) +
        scale_x_discrete(labels = c("unst", "LPS2"))
    print(current_plot)
}
```

We see that there are only two equivalence classes for this gene, `58069` and
`58070`. In addition, these equivalence classes are of length 1. This means
that each equivalence class corresponds only to a single transcript (isoform).

```{r}
rowData(sce)[c("58069","58070"),"tx_id"]
```

As such, even though we performed an analysis on the equivalence class level,
we here can attribute the observed change in usage to specific transcripts! 
I.e., we can say that both transcript "ENSMUST00000032779.12" and transcript
"ENSMUST00000131108.9" are beeing differentially used between unstimulated
and LPS (2 hours) stimulated cells.

Below, we show an example of a geene for which the biological interpretation 
will be more ambiguous.

### Biological interpretation for gene ENSMUSG00000058427.11

```{r}
plots <- satuRn::plotDTU(
    object = sce,
    contrast = "Contrast1",
    groups = list(group1, group2),
    coefficients = list(c(1, 0), c(0, 1)),
    summaryStat = "model",
    transcripts = NULL,
    genes = "ENSMUSG00000006818.6",
    top.n = 3)

# to have same layout as in our publication
for (i in seq_along(plots)) {
    current_plot <- plots[[i]] +
        scale_fill_manual(labels = c("unst", "LPS2"), 
                          values = c("royalblue4","firebrick")) +
        scale_x_discrete(labels = c("unst", "LPS2"))
    print(current_plot)
}
```

Gene "ENSMUSG00000058427.11" has 3 equivalence classes: `119653|119657`, 
`119653|119656|119657` and `119653`. Only equivalence class 
`119653|119656|119657` is flagged as significantly differentially used between
the two treatment groups. However, when we look at the feature-level 
information,

```{r}
rowData(sce)[c("119653", "119653|119657", "119653|119656|119657"),"tx_id"]
```

we see that this equivalence class contains reads that are compatible with both
transcript "ENSMUST00000007012.6" and "ENSMUST00000233897.2". As such, we cannot
attribute the observed difference in usage to a specific transcript.

Furthermore, transcript "ENSMUST00000007012.6" is involved in all equivalence
classes, and transcript "ENSMUST00000233897.2" is involved in both the 2nd
and third equivalence class. Equivalence class 1 and 3 are both not 
significantly differentially used. As such, among the three transcripts involved
in this gene, we did not get a clear picture which of the specific isoforms
are differentially used.

# Conclusions

In conclusion, we here displayed how to perform a equivalence class-level 
differential transcript usage analysis using `satuRn`. The biggest advantage of
equivalence classes is that they can even be obtained from sequencing protocols 
that do not allow for transcript level quantification, which includes many of 
the droplet-based protocols such as 10X Genomics Chromium.  Their biggest 
drawback is their downstream biological interpretation. In some cases, it will
be possible to uniquely link a equivalence class to a specific isoform. However,
in most cases, multiple transcripts can be compatible with the equivalence 
class. In those cases, the interpretation is often limited to recognizing that
something is changing internally in the gene, but its unclear which exact 
isoforms are involved. This, however, still provides sub gene-level information,
and may prompt follow-up experiments with a different methodolgy that allows for
pinpointing the exact transcript-level change.

# Session info

```{r}
sessionInfo()
```

# References









